---
title: "Using the ModSeqR package"
author: "Hailey Zimmerman and Jonathon T. Hill, PhD"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    css: "bioconductor2.css"
vignette: >
  %\VignetteIndexEntry{Using the ModSeqR package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---
  
```{r, include=FALSE}
library(knitr, quietly=TRUE)
library(ModSeqR, quietly=TRUE)
# library(Biostrings, quietly=TRUE)
opts_chunk$set(tidy=FALSE, tidy.opts=list(width.cutoff=70))
```

# Introduction

The `ModSeqR` package provides basic functions for importing and 
working with third-generation methylation sequencing data.

This vignette will walk you through a typical workflow. If sequencing data is coming from
Wasatch Biolabs (https://www.wasatchbiolabs.com/), `.ch3` files will be delivered directly if requested
for your batch. These `.ch3` files contain methylation calls from a third-generation sequencing run. 
A `.ch3` file will typically be created for each sample within a run, and is meant 
as an intermediate existing to compress large amounts of data into a usable form.

If not delivered through Wasatch Biolabs, `.ch3` files can be built using this package.

It's most helpful to store all the sample CH3 files from a run you want to 
analyze within a single directory to get started. 

# The mod_db object and piping

Most ModSeqR functions return an object of class `mod_db` — a lightweight list that holds four
pieces of state:

| Slot | Contents |
|---|---|
| `db_file` | Path to the `.mod.db` DuckDB file on disk |
| `current_table` | Name of the most recently written table (updated automatically) |
| `con` | Active DBI connection while a function runs; `"none"` when closed |
| `last_result` | The most recently computed in-memory result (see below) |

Typing a `mod_db` variable — or calling `print()` on it — shows all four slots:

```{r eval=FALSE}
print(mod_db)
# <mod_db object>
# Database file:
#   my_data.mod.db
# Current table:
#   mod_diff_windows
# Connection:
#   NULL
# Last result: <matrix> [6 x 6]
```

## Piping

Because every pipeline function returns `invisible(mod_db)`, you can chain them with `|>`:

```{r eval=FALSE}
mod_db <- make_mod_db(ch3_files, "my_db") |>
  summarize_mod_windows()                 |>
  calc_mod_diff(cases = c("A","B"), controls = c("C","D"))
```

## Capturing in-memory results with `get_mod_result()`

Some functions compute a result that is printed to the console but not otherwise accessible.
These functions store the result in `mod_db$last_result` so you can retrieve it after the pipe
without breaking the chain:

| Function | What `get_mod_result()` returns |
|---|---|
| `calc_mod_samplecor()` | Numeric correlation `matrix` |
| `plot_mod_pca()` | `list(pca = <prcomp object>, scores = <data frame>)` |
| `calc_mod_diff()` | `data.frame` preview (head) of the result table |

```{r eval=FALSE}
# Correlation matrix
mod_db  <- calc_mod_samplecor(mod_db, call_type = "windows")
cor_mat <- get_mod_result(mod_db)      # numeric matrix
corrplot::corrplot(cor_mat)            # pass to any downstream function

# PCA
mod_db    <- plot_mod_pca(mod_db)
pca_list  <- get_mod_result(mod_db)
pca_list$scores                        # data frame of PC coordinates per sample
biplot(pca_list$pca)                   # use the prcomp object directly

# Differential methylation preview
mod_db  <- calc_mod_diff(mod_db, cases = ..., controls = ...)
preview <- get_mod_result(mod_db)      # head() of the result table
full_df <- get_mod_table(mod_db, mod_db$current_table)  # full table
```

# Creating `.ch3` files from a TSV (optional)

If your methylation calls are in a delimited text file (e.g., from a `modkit extract-calls` TSV),
you can convert them into compressed `.ch3` archives using `make_mod_archive()`. 
This will dramatically reduce file size and allow use of the ModSeqR package.

`make_mod_archive()` reads a tab-delimited calls file, selects and transforms the 
required columns, and writes a compressed Arrow dataset to `out_path` using zstd compression. 
The function derives `start`/`end` coordinates per CpG (accounting for strand and the CG k-mer), 
optionally shortens long `read_id`'s to shrink file size (`short_ids = TRUE`), and embeds 
your `sample_name` into the output filenames. The output may be sharded into multiple 
files (e.g., `SampleA-0.ch3`, `SampleA-1.ch3`, …) depending on Arrow’s writer.

## Coordinate convention:
The `.ch3` files use 0-based, half-open coordinates for `start`/`end` (like BED format). This means the `start` column is inclusive and `end` is exclusive. Keep this in mind if you compare `.ch3` positions to genome browsers or annotation files that use 1-based inclusive coordinates (e.g., GTF/GFF). Conversion may be required when joining with external annotations.

## Expected input columns (typical from modkit):
`read_id`,`forward_read_position`, `ref_position`, `ref_mod_strand`, `chrom`,
`read_length`, `query_kmer`, `call_prob`, `call_code`, `base_qual`, `flag`.

After the `.ch3` files are written, point `make_mod_db()` at the directory containing them (or pass their paths directly), and the package will take care of the rest.

```{r eval=FALSE}
# Convert a modkit TSV to a compressed .ch3 archive
make_mod_archive(
  file_name   = "/path/to/modkit_calls.tsv",  # tab-delimited input
  sample_name = "Astrocytes",                 # used in output filenames
  out_path    = "/path/to/ch3_out",           # directory to write .ch3 files
  short_ids   = TRUE                          # shorten read_id to reduce size
)

# The directory now contains e.g.: Astrocytes-0.ch3, Astrocytes-1.ch3, ...
# Build a DB from those .ch3 files:
make_mod_db(
  ch3_files = "/path/to/ch3_out",   # a directory OR a vector of file paths
  db_name   = "astro_db"
)
```

## Notes & tips
* If your input file contains unplaced minus-strand rows at position 0 (common in some extract outputs), those are filtered automatically.
* If you already have concise read IDs, set short_ids = FALSE.
* You can run make_mod_archive() for each sample TSV, all writing into the same out_path; make_mod_db() can then ingest the whole directory in one call.
* The .ch3 files are standard Parquet—fast to read, compact on disk, and splittable for parallel import.

# Loading Data
The first step of a workflow will be to upload data from a directory holding 
sample sequencing data in the form of CH3 files. These files are typically 
created by Wasatch Biolabs.

Once data is obtained in this format, the first step is done using the 
`make_mod_db()` function. This function directly imports a directory of CH3 
files, and creates a database in the form of a local file.

## make_mod_db
`make_mod_db()` builds the DuckDB database from one or more `.ch3` Parquet files. 

It requires two arguments: `ch3_files`, which can be a directory or a vector of file paths, and `db_name`, which is the path where the DuckDB file will be created (the .mod.db suffix is added if missing). 

You may optionally pass a named character vector to assign explicit `sample_name`s; otherwise, `sample_name` is derived from the file’s basename by removing a trailing -<digits> (if present) and the .ch3 extension (e.g., Astrocytes-12.ch3 → Astrocytes, M1.combined.ch3 → M1.combined). Additional filters include `chrom` (vector of chromosomes to keep), `min_read_length` (default 50), `min_call_prob` (default 0.9), `min_base_qual` (default 10), and an optional `flag` filter. The function configures DuckDB pragmas for performance (temp directory, threads ≈ all-but-one core, memory ≈ 50% RAM), drops any existing tables, and creates a `calls` table with: `sample_name`, `chrom`, `start`, `end`,`read_position`, `call_code`, `read_length`, `call_prob`, `base_qual`, and `flag`.

Here is an example:

```{r eval=FALSE}
# Example: directory of CH3 files; auto-derived sample names
ch3_files <- system.file("TEST_DATA", package = "ModSeqR")   # adjust as needed
mod_db    <- tempfile("example_db")
make_mod_db(ch3_files, mod_db)

# Example: explicit sample names override the fallback
make_mod_db(
  ch3_files = c(
    sample1  = "/data/sample1-0.ch3",
    sample2     = "/data/sample2.ch3"
  ),
  db_name   = "named_samples"
)
```


After this step, a database is made and the raw calls table is included. A user 
may continue their analysis in any direction they want. They may run a quality control 
(`run_mod_qc()`), summarize data by positions (`summarize_mod_positions()`),
regions (`summarize_mod_regions()`), or windows (`summarize_mod_windows()`). 
At any point in the analysis, a user can export a table in the database by calling
`export_mod_tables()`, or use `get_mod_table()`to import as a tibble into the local environment.
All these functions will be explained below.

## summarize_mod_positions
This function summarizes methylation data (from the calls table) by each unique 
position in the genome. A table called "positions" will be built into the database.

`summarize_mod_positions()` aggregates the `calls` table at each genomic position for each sample and writes a `positions` table. It always produces `num_calls`, and, for each unique modification you request, a `<label>_counts` and `<label>_frac` column (ex. methylation would be m_counts and m_frac). You control modification types via `mod_code` (e.g., c("m","h","m + h") or a novel "a"). 

Labels are formed by removing spaces/“+” (so `"m + h"` becomes `mh`). The unmodified class is defined by `unmod_code` (default `"-"`) and named by `unmod_label` (default `"c"`, yielding `c_counts` and `c_frac`). Optional `chrs` (defaults to all chromosomes in the data), `samples`, and `min_num_calls` let you filter chromosomes, samples, and minimum coverage per position.

```{r eval=FALSE}
# Minimal usage: default m/h classes
summarize_mod_positions(mod_db)

# Custom classes, including a novel code 'a', with stricter filtering
summarize_mod_positions(
  mod_db,
  mod_code      = c("a", "m + h"),
  unmod_code    = "-",
  unmod_label   = "c",
  min_num_calls = 5
)
```

## summarize_mod_regions
`summarize_mod_regions()` joins per-position counts to an annotation file of regions (BED/TSV/CSV with `chrom`, `start`, `end`, and optional `region_name`) and writes a `regions` table. If `region_name` is missing, it is synthesized as `chrom_start_end`. 

You choose the join type via `join = "inner" | "left" | "right"`; inner keeps only overlapping positions/regions, left keeps all regions (filling zeros where appropriate), and right keeps all positions with region matches. 

As with positions, you specify any set of `mod_code` values (single or combined like "m + h"), `unmod_code`, `unmod_label`, and filters (`chrs`, `samples`, `min_num_calls`). Output columns include `sample_name`, `region_name`, `chrom`, `start`, `end`, `num_sites` (number of unique positions aggregated), `num_calls`, and the `<label>_counts/<label>_frac` columns for all requested labels.

```{r eval=FALSE}
region_bed = system.file("Islands_hg38_test.csv", package = "ModSeqR")
# Minimal call
summarize_mod_regions(mod_db, region_file = region_bed)

# Custom call
summarize_mod_regions(
  mod_db,
  region_file   = region_bed,
  join          = "inner",
  mod_code      = c("m","h","m + h"),   # supports novel codes too, e.g., "a"
  min_num_calls = 5
)
```

## summarize_mod_windows
`summarize_mod_windows()` produces sliding/offset windows from per-position counts and writes a windows table. 

By default it tiles with 1,000 bp windows and 10 bp offsets so positions are assigned to multiple staggered windows. The output contains `num_sites`, `num_calls`, and `<label>_counts/<label>_frac` for all requested modification labels (as defined by `mod_code`, `unmod_code`, and `unmod_label`). 

You can adjust `window_size`, `step_size`, `min_num_calls`, limit to particular `chrs` (defaults to all chromosomes in the data) or `samples`, and choose to `overwrite` an existing table. Default overwrite is TRUE.

```{r eval=FALSE}
# 100 bp windows, non-overlapping (step_size = window_size)
summarize_mod_windows(mod_db, window_size = 100, step_size = 100)

# 2 kb windows with offsets; include a novel 'a' class
summarize_mod_windows(
  mod_db,
  window_size   = 2000,
  step_size     = 20,
  mod_code      = c("a", "m + h"),
  min_num_calls = 25,
  overwrite     = TRUE
)
```

## summarize_mod_reads
This function summarizes methylation data from a mod_db database by the reads.
A table called "reads" will be built into the database.

```{r eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Calculate windows
summarize_mod_reads(mod_db)
```

There is only 1 required arguments. `mod_db` is a list containing the database 
file path. This should be a string representing a path to the database, or a valid
"mod_db" class object. Optionally, `key_table` is a path to an external table which
can be used to filter for unique windows of interest. It should have a "chrom", 
"start", and "end" column to specify which windows are of interest. The reads table 
will be filtered for just those windows from the key_table.Lastly, `min_CGs` specifies
the minimum number of CG sites for a read to be included. If a read has less than 
that argument, it will not be included in the table.

# Convenience utilities (inspect/rename/tidy)

Sometimes you just want to peek at what’s in your database, check a table’s
shape, or quickly fix sample names. The helpers below are lightweight wrappers
around DuckDB/DBI that do exactly that.

## get_mod_stats()

Before or after building a database, you can compute summary statistics directly
from a `.ch3` file using `get_mod_stats()`. This operates on the raw Parquet file
(not the database) and reports total calls, unique reads, mean read length, CpG
coverage at several minimum-read thresholds, flag distributions, and high-confidence
call rates.

Results are printed to the console. Because the function returns the list
*invisibly*, you can also capture it for programmatic use:

```{r eval=FALSE}
# Print stats to console
get_mod_stats("sample1.ch3")

# Capture for downstream use
stats <- get_mod_stats("sample1.ch3")
stats$num_reads
stats$cpg_coverage

# Write to a log file instead of the console
get_mod_stats("sample1.ch3", log_file = "sample1_stats.log", silently = TRUE)
```

## get_mod_dbinfo()

Prints a quick summary of the database: on-disk size, a list of tables, and the
distinct sample_names (if a calls table exists). Invisibly returns the
closed mod_db object.

```{r eval=FALSE}
# Path or mod_db object both work
get_mod_dbinfo("my_data.mod.db")
```

## get_mod_tableinfo()

Summarizes a single table: total rows, per-sample row counts (if the table has a
sample_name column), and a list of columns. Good first check after creating a
table.

```{r eval=FALSE}
get_mod_tableinfo("my_data.mod.db", table_name = "positions")
```

## get_mod_cols()

Returns (and prints) the column names for a given table. Handy when you’re about
to write a query and can’t remember exact column names.

```{r eval=FALSE}
get_mod_cols("my_data.mod.db", "windows")
```

## get_mod_cpg_count()

Counts unique CpG sites in a table, where uniqueness is defined by distinct
(start, end) pairs. Prints a small summary and invisibly returns the count.

```{r eval=FALSE}
n_cpg <- get_mod_cpg_count("my_data.mod.db", table_name = "calls")
```

## rename_mod_samples()

Renames sample_names directly in a specified table (e.g., positions,
windows, regions, calls). You provide a mapping from old → new names as a
named character vector (c(old = "new", ...)) or a two-column data frame
with old and new. The function validates inputs, updates in place, and
(optionally) prints a before/after list of distinct names.

```{r eval=FALSE}
# Rename in the 'positions' table
rename_mod_samples(
  "my_data.mod.db",
  table       = "positions",
  samples_map = c("Cortical_Neurons" = "Cortex", "Blood" = "PBMC"),
  preview     = TRUE
)
```

## remove_mod_table()

Deletes a table by name if it exists, with a clear message either way (no error
if the table is missing). Useful to clean up experiments.

```{r eval=FALSE}
remove_mod_table("my_data.mod.db", "tmp_debug_table")
```


*Tip*: All helpers accept either a path to the *.mod.db file or an existing
mod_db object. They open a connection, do the work, and close it for you.*

# Differential Methylation
## calc_mod_diff
This function calculates differential methylation between specified case and control groups using various statistical methods. 

Differential methylation can be calculated from <b><i>positional</i></b>, <b><i>regional</i></b>, or <b><i>window</i></b> methylation data.

Calling this function creates a unique `mod_diff_*` table within the database, where differential methylation results are stored.

The table name depends on the input `call_type`:

* `call_type = "positions"` → `mod_diff_positions`

* `call_type = "regions"` → `mod_diff_regions`

* `call_type = "windows"` → `mod_diff_windows`

Calling `calc_mod_diff()` again on the same type of data will overwrite the existing `mod_diff_{type}` table (e.g. `mod_diff_windows` will be rewritten if you run it again on window data).

```{r eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
  
# Get methylation statistics for the 'positions' call type without plotting
calc_mod_diff(mod_db = mod_db,
              call_type = "positions",
              cases = "blood",
              controls = "sperm")
``` 

### Modification type (mod_type)

The `mod_type` argument specifies which modification to test. The default is `"mh"` (methylation/hydroxymethylation). Other codes include `"a"` for 6mA, `"17596"` for inosine, and `"17802"` for pseudouridine. Bare numeric codes are automatically prefixed with `"m_"`, so `mod_type = "17596"` and `mod_type = "m_17596"` are equivalent.


### Automatic choice of statistical test (calc_type)

The argument `calc_type` controls which statistical method is used to test for differential methylation.
It can be set manually to one of:

* `"beta_bin"` – beta-binomial likelihood-ratio test (accounts for overdispersion across replicates)

* `"wilcox"` – Wilcoxon rank-sum test on per-sample methylation fractions

* `"fast_fisher"` – fast Fisher's exact test on aggregated counts

* `"r_fisher"` – Fisher's exact test via base R

* `"log_reg"` – logistic regression–based test


If `calc_type` is left as `NULL` (the default), the function automatically chooses an appropriate method based on how many samples are in each group:

* **5 or more samples in both groups** → `"wilcox"` — enough samples for a non-parametric rank-based test.

* **Fewer than 5 samples in either group** → `"fast_fisher"` — directly compares aggregated counts.

A message is printed indicating which method was chosen. You can always override the automatic selection by explicitly setting `calc_type` yourself — for example, to use `"beta_bin"` with 2–4 replicates per group (see *Beta-binomial test details* below).

### Why this behavior?

The choice of statistical test depends on how many biological replicates are available:

* **With many replicates (≥ 5 per group)**, the Wilcoxon rank-sum test works well. It compares the distribution of methylation fractions between groups without assuming a specific distribution shape. With enough samples, rank-based tests have good power and produce granular p-values.

* **With no or few replicates (< 5 per group)**, the fast Fisher's exact test is used by default. There is limited between-sample variation to model, and the test directly compares modification counts in a 2×2 contingency table. For 2–4 replicates where biological overdispersion is a concern, consider explicitly setting `calc_type = "beta_bin"` (see below).

### Beta-binomial test details

The beta-binomial test (`calc_type = "beta_bin"`) models each sample's modified-read count at a locus as a draw from a beta-binomial distribution. This captures two sources of variation: sampling noise (from finite read counts) and biological variability (different samples having genuinely different modification rates at the same site).

The test fits two models at each locus:

* **Null model:** all samples (case + control) share one mean modification rate μ
* **Alternative model:** cases have μ~case~ and controls have μ~control~, with shared overdispersion

The difference in log-likelihoods is compared to a χ² distribution to obtain a p-value. The output includes an `overdispersion` column (φ, ranging 0–1) that indicates how much the true modification rate varies across replicates at each locus. Values near 0 mean replicates are very consistent; higher values indicate more biological variability.

Key advantages of the beta-binomial:

* Samples with higher coverage naturally contribute more to the estimate (a sample with 500 calls is more informative than one with 10 calls)
* Biological variability between replicates is explicitly modeled, not ignored
* No additional R package dependencies are required (implemented in base R)

**Performance note:** The beta-binomial test runs `optim()` independently at every locus, which makes it substantially slower than the vectorized Fisher or Wilcoxon tests. For a typical whole-genome window analysis, expect run times on the order of **minutes to hours** depending on the number of windows, step size, and number of samples. To manage run times:

* Use larger `window_size` and `step_size` in `summarize_mod_windows()` to reduce the total number of windows
* Apply `min_sites` filtering to drop low-quality windows before testing
* Use `threads` and `memory_limit` parameters to control resource usage

```{r eval=FALSE}
# Auto-select: >= 5 per group → wilcox
calc_mod_diff(mod_db,
              call_type = "windows",
              cases    = c("case1", "case2", "case3", "case4", "case5"),
              controls = c("ctrl1", "ctrl2", "ctrl3", "ctrl4", "ctrl5"))

# Auto-select: < 5 per group → fast_fisher
calc_mod_diff(mod_db,
              call_type = "windows",
              cases    = c("case1", "case2", "case3"),
              controls = c("ctrl1", "ctrl2", "ctrl3"))

# Explicitly request beta-binomial for 2–4 replicates
calc_mod_diff(mod_db,
              call_type = "windows",
              cases     = c("case1", "case2", "case3"),
              controls  = c("ctrl1", "ctrl2", "ctrl3"),
              calc_type = "beta_bin")
```

### Site coverage filtering with `min_sites`

When working with window-level data, some windows may have very few
modification sites covered in certain samples. The `min_sites` parameter
filters these out before testing.

`min_sites` is an integer representing the minimum number of distinct
positions within a window that must have modification calls for each
sample. For CpG methylation, this is the number of CpGs covered. Windows
where any sample falls below this threshold are dropped before statistical
testing.

```{r eval=FALSE}
# Require at least 20 CpGs covered per sample per window
calc_mod_diff(mod_db,
              call_type = "windows",
              cases = c("sample1", "sample2", "sample3"),
              controls = c("ctrl1", "ctrl2", "ctrl3"),
              min_sites = 20)
```

**Guidance on values:**

* `min_sites = 5` — lenient; filters only the most sparse windows
* `min_sites = 10–20` — moderate; good default for most analyses
* `min_sites = 50+` — stringent; for large windows with high coverage

Note: `min_sites` only applies when the input table has a `num_sites`
column (i.e., window or region data). It is silently skipped for
position-level analyses. If the filter removes all windows, the function
will stop with a message suggesting you lower the threshold.


## collapse_mod_windows
This function only works on a mod_diff_windows table already in the database.
After a differential methylation analysis is called on windows, this will collapse 
significant windows by merging contiguous regions that meet the specified criteria.

The only required parameter is the link to the database. Optionally, a user may specify
the following arguments.`max_distance`is the maximum allowable distance between consecutive
significant windows for merging (default: 1000). `sig_cutoff` is the significance 
threshold for adjusted p-values (default: 0.05). Lastly, `min_diff` specifies the
minimum absolute methylation difference required for inclusion in the analysis (default: 0.5).

```{r, eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
  
# Collapse significant windows
collapse_mod_windows("my_data.mod.db")
``` 

## run_mod_analysis
This function automates an entire workflow in one command. It identifies differences in DNA methylation patterns between two groups (cases and controls) by summarizing methylation data at various genomic scales (positions, regions, or windows) and then performing statistical tests, outputting the results into organized CSV files within one directory.

`run_mod_analysis()` creates a new subdirectory named "Mod_Diff_Analysis_Results" within the out_path where all the data will be written to. It writes three CSV files into the "Mod_Diff_Analysis_Results" directory by directly copying tables from the DuckDB database:

**mod_diff.csv**: Contains the results of the differential analysis (e.g., p-values, fold changes) for each position, region, or window. The source table in the database is determined by diff_table_name.

**All_CpGs.csv**: Contains all the summarized methylation data for the chosen call_type (positions, regions, or windows) across all samples. The source table in the database is determined by call_type.

**Sig_CpGs.csv**: Contains summarized methylation data for only those positions/regions/windows that are deemed significantly differentially methylated, based on the p_val_max threshold.

```{r, eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
  
# Collapse significant windows
run_mod_analysis(mod_db, 
             out_path = "/Users/analysis",
             call_type = "windows",
             cases = c("sperm"),
             controls = c("blood"))
``` 

# Quality Control 

A variety of quality control functions are available to visually assess 
methylation data. 

## plot_mod_cov
This function calculates and optionally plots statistics for coverage data from
methylation sequencing experiments. It can handle both positional and regional
methylation data.

```{r eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Get coverage statistics for the 'positions' call type
plot_mod_cov(mod_db = mod_db, call_type = "positions")
``` 

## plot_mod_modfrac
This function retrieves and calculates methylation statistics (mean methylation 
fractions) from a specified table in the mod database. It can either return 
summary statistics or plot a histogram of the methylation values, depending on 
the user's preference. 

```{r eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Get methylation statistics for the 'positions' call type without plotting
plot_mod_modfrac(mod_db = mod_db, call_type = "positions")
``` 

## calc_mod_samplecor
This function calculates and optionally plots a correlation matrix for
methylation or other modification fraction data from genomic positions.
It can handle both position-based and region-based calls and supports
visualization using ggplot2.

The correlation matrix is stored in `mod_db$last_result` and can be retrieved
with `get_mod_result()` after the call.

```{r eval=FALSE}
mod_db  <- calc_mod_samplecor(mod_db = mod_db, call_type = "positions", plot = TRUE)

# Capture the correlation matrix for downstream use
cor_mat <- get_mod_result(mod_db)
```

## plot_mod_pca
This function performs Principal Component Analysis (PCA) on methylation data
retrieved from the database. It aggregates the methylation fraction data
based on the specified call type and prepares it for PCA analysis.

The PCA result is stored in `mod_db$last_result` as a list with elements `pca`
(the `prcomp` object) and `scores` (a data frame of PC coordinates per sample).

```{r eval=FALSE}
mod_db   <- plot_mod_pca(mod_db)

# Capture the PCA result for downstream use
pca_list <- get_mod_result(mod_db)
pca_list$scores   # PC coordinates per sample
``` 


## run_mod_qc

Combining all functions above, the `run_mod_qc()`` function computes all QC checks
on methylation data at the same time. 

```{r eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Calculate windows
run_mod_qc(mod_db)
``` 
 
# Retrieving and Exporting Tables
You can access data in your database with two different functions.

## get_mod_table
To retrieve a single table into your local environment for further analysis within R, 
using `get_mod_table` will transform it into a tibble and store it into a variable.
Arguments required includes the database, and name of the table to be retrieved.

```{r, eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Get any table in your database and create a variable in R to work with
positions = get_mod_table("my_data.mod.db", "positions")
regions = get_mod_table("my_data.mod.db", "regions")

mod_diff_regions = get_mod_table("my_data.mod.db", "mod_diff_regions")
collapsed_windows = get_mod_table("my_data.mod.db", "collapsed_windows")
```

## export_mod_table
To export one or more tables into a CSV file, use `export_mod_table`. This function 
takes three arguments, the database, which tables to export, and the path where 
the CSV files will be saved. 

```{r, eval=FALSE}
# Specify the path to the database
mod_db <- system.file("my_data.mod.db", package = "ModSeqR")
 
# Export the table to your computer
export_mod_table("my_data.mod.db", 
             table = "windows", 
             out_path = "/Desktop/My_Folder/windows.csv")

export_mod_table("my_data.mod.db", 
             table = "mod_diff_windows", 
             out_path = "/Desktop/My_Folder/mod_diff_windows.csv")
```


# Conclusion
In this vignette, we have walked you through the basic functions in the 
`ModSeqR` package. This work is a work in progress and we hope to 
improve its functionality. If you have any suggestions or 
requested features, please email Jonathon Hill at jhill@byu.edu.
